### 2021-10-26 documents/sections/extensions.tex
regular verbs in data and docs
added more regular verbs

\subsection{\PPek \rc{k-}}
\label{sec:pekodian}
The Pekodian branch was suggested by \textcite{meira2005southern}, as the result of fieldwork on \bakairi by Meira and the availability of more material on \ikpeng. % \parencites{pacheco1997relativizacao}{ikpengpacheco1997}{campetela1997analise}{pacheco1998nominalizacao}{ikpengpacheco2001}{campetela2002aspectos}{pacheco2002aspectos}{pacheco2003intransitivos}.
It consists of closely related \arara and \ikpeng, with \bakairi as a more distant member.
\textcite{meira2005southern} focused on phonological and lexical properties, so no reconstructive work has been done on \PPek morphosyntax.
All three Pekodian languages have a regular \gl{1}\gl{s_a_} marker \obj{k-}, as evidenced by the paradigms in \exref{pekreg}.
Its presence can thus be reconstructed to \PPek.

\ex<pekreg>
\begin{tabular}[t]{@{}llll@{}}
& \bakairi \qu{to go up}  & \arara \qu{to dance}  & \ikpeng \qu{to run} \\
& \parencite[4]{meira2003bakairi} & \parencite[150]{alves2017arara} &  \parencite[52]{ikpengpacheco2001}\\
\gl{1}\gl{s_a_} & \obj{\emp{k-}əku-} & \obj{\emp{k-}origu-} & \obj{\emp{k-}aranme-} \\
\gl{2}\gl{s_a_} & \obj{m-əku-} & \obj{m-origu-} & \obj{m-aranme-} \\
\gl{1+2}\gl{s_a_} & \obj{kɨd-əku-} & \obj{kud-origu-} & \obj{kw-aranme-} \\
\gl{3}\gl{s_a_} & \obj{n-əku-} & ∅\obj{-origu-} & ∅\obj{-aranme-} \\
\end{tabular}
\xe

@@

The V-initial nature of \qu{to go} is evidenced in its Xinguan forms, while \bakairi \phon{\rc{wɨ}}{\obj{u}} resulted in a shift of the morpheme boundary, see also \cref{sec:go}.
The \ikpeng form \obj{aran} is made compatible with our reconstruction \rc{ɨtən} by considering that \ikpeng \obj{a} is an attested outcome of \rc{ə} (\PXin \rc{o}/\obj{e}), for example \begin{inlinelist}
 \item \obj{akari} \qu{dog} in \cref{tab:pxinw} above
 \item \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg
 \item \obj{jaj} \qu{tree} \parencite[98]{ikpengpacheco2001} from \PC \rc{jəje}
 \end{inlinelist}.
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{\underline{ə}t\underline{ə}n}: \obj{\underline{e}r\underline{o}-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

@@

\subsection{\PWai \rc{k-}}
\label{sec:waiwaian}
This extension was briefly discussed in \cref{sec:extensions_intro}; the new \gl{1}\gl{s_p_} prefix \rc{k-}, already innovated at the \PPar stage, was later extended to \gl{1}\gl{s_a_} in \PWai.
This created a unified \gl{1}\gl{s} category, reflected in both \hixka and \waiwai \exref{wais}.

\ex<wais> Waiwaian \rc{eɸurka} \qu{to fall} (\gl{s_a_}) and \rc{wɨnɨkɨ} \qu{to sleep} (\gl{s_p_})\\
\begin{tabular}[t]{@{}lllllll@{}}
& \multicolumn{2}{l}{\PWai} & \multicolumn{2}{l}{\hixka}  & \multicolumn{2}{l}{\waiwai} \\
 & \gl{s_a_} & \gl{s_p_} & \gl{s_a_} & \gl{s_p_} & \multicolumn{2}{l}{\gl{s}} \\
\gl{1} & \rc{\emp{k-}eɸurka-} & \rc{kɨ-wɨnɨkɨ-} & \obj{\emp{k-}ehurka-} & \obj{kɨ-nɨkɨ-} & \obj{\emp{k-}eɸɨrka-} & \obj{kɨ-wɨnk-} \\
\gl{2} & \rc{m-eɸurka-} & \rc{o-wɨnɨkɨ-} & \obj{m-ehurka-} & \obj{o-wnɨkɨ-} & \obj{m-eɸɨrka-} & \obj{mɨ-wɨnk-} \\
\gl{1+2} & \rc{t-eɸurka-} & \rc{tɨt-wɨnɨkɨ-} & \obj{t-ehurka-} & \obj{tɨ-nɨkɨ-} & \obj{t͡ʃ-eɸɨrka-} & \obj{tɨt-wɨnk-} \\
\gl{3} & \rc{ɲ-eɸurka-} & \rc{nɨ-wɨnɨkɨ-} & \obj{ɲ-ehurka-} & \obj{nɨ-nɨkɨ-} & \obj{ɲ-eɸɨrka-} & \obj{nɨ-wɨnk-} \\
\end{tabular}\\
\parencites[510]{howard2001wrought}[189--191]{hixkaryanaderby1985}[209--211]{hawkins1953waiwai}[50]{waiwaihawkins1998}
\xe
%
\waiwai \obj{ka} \qu{to say} does not take \obj{kɨ-}, but rather conservative \obj{wɨ-} \exref{hixɨ.wai-108}.
Its \hixka counterpart also takes a reflex of \gl{1}\gl{s_a_} \rc{w-}, which has the form \obj{ɨ-} \exref{hixɨ.hix-118}.
A formally identical prefix is found on transitive verbs in \gl{1}>\gl{3} scenarios in \hixka \exref{hixɨ.hix-17}, where it also corresponds to \waiwai \obj{w(ɨ)-} \exref{hixɨ.wai-159}.

@@

\subsection{\PTir \rc{t-}}
\label{sec:taranoan}
The Taranoan languages were already grouped together by \textcite{girard1971proto}, consisting of closely related \trio and \akuriyo (subsumed here under the moniker Tiriyoan), and the more distant member \carijo.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar.
The \setone paradigms of \trio and \akuriyo contain an interesting puzzle: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.
%This is illustrated by comparing reconstructible \PTir forms with more conservative \wayana \exref{sat} and \kalina \exref{13w}, which can be taken as representative for \PC.
%
%\ex<sat>
%\begin{tabular}[t]{@{}llllll@{}}
%\gl{1}\gl{s_a_} & \PC & \wayana & \PTir & \trio & \akuriyo\\
%\qu{yawn} & \rc{w-e-mɨta-po-} & \obj{w-emtapɨ-} & \rc{t͡ʃ-entapo-} & \obj{s-entapo-} & \obj{t͡ʃ-entapo-}\\
%\qu{bathe} & \rc{w-e-pɨ-} & \obj{w-epɨ-} & \rc{t͡ʃ-epɨ-} & \obj{s-epɨ-} & \obj{t͡ʃ-epɨ-}\\
%\qu{talk} & \rc{w-ət-uru} & \obj{w-əturu-} & \rc{t-əturu-} & \obj{t-əturu-} & \obj{k-əturu-}\footnotemark\\
%\qu{learn} & \rc{w-ət-eme-pa} & \obj{w-əhepa-} & \rc{t-əempa-}\footnotemark & \obj{t-əenpa-} & \obj{k-əempa-}\\
%\end{tabular}\\
%\parencites[28]{camargo2010wayana}[292, 664]{triomeira1999}[77, 86, 82]{gildea1994akuriyo}[50, 208, 213]{wayanatavares2005}
%\xe
%\addtocounter{footnote}{-1}
%\footnotetext{\PTir \rc{t-} was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.}
%\addtocounter{footnote}{1}
%\footnotetext{
%The \trio form is not due to dissimilation from \rc{mp} to \obj{np}, but rather a phonemic transcription rendering the syllable-final archiphoneme \pnm{N} as \obj{n} \parencite[31]{triomeira1999}.
%Further, \PTar intervocalic \rc{t͡ʃ} was lost entirely in \trio (except after \rc{i}) and is reflected as \obj{ʔ} in \akuriyo \parencite[31--32]{meira1998proto}.
%Given the original shape of the detransitivizers \rc{ət(e)-} and \rc{e-}, \trio \obj{ə-enpa} \qu{to learn} contains a reflex of the former with loss of \rc{t}, so \akuriyo \obj{əempa} would be expected to have a glottal stop: ?\obj{əʔempa}.
%Its absence could either be due to \trio influence or not being transcribed by \textcite{gildea1994akuriyo}.
%Since there is no evidence for a glottal stop in \PTir (\rc{əʔempa}), I do not reconstruct it.}
%
%\ex<13w> 
%\begin{tabular}[t]{@{}llllll@{}}
%\gl{1}>\gl{3}& \PC & \kalina & \PTir & \trio & \akuriyo \\
%\qu{see} & \rc{t-ene-} & \obj{s-ene-} & \rc{w-ene-} & \obj{w-ene-} & \obj{(w-)ene-} \\
%\qu{bring} & \rc{t-enepɨ-} & \obj{s-enepɨ-} & \rc{w-enepɨ-} & \obj{w-enepɨ-} & Ø\obj{-enepɨ-} \\
%\end{tabular}\\
%\parencites[425--426]{hoff1968carib}[28, 304]{triomeira1999}[86--87]{gildea1994akuriyo}
%\xe
%%
%Thus, \wayana preserves the \PC \gl{1}\gl{s_a_} marker \rc{w-}, while it was replaced by \rc{t-} (\phonr{}{\rc{t͡ʃ-}}{\rc{e}}) in \PTir \exref{sat}.
%%Both \qu{to yawn} and \qu{to bathe} are transparently derived \gl{s_a_} verbs, from \rc{mɨta-po} \qu{to open the mouth of} and \rc{pɨ} \qu{to wash}, respectively.
%On the other hand, \gl{1}>\gl{3} scenarios show a reflex of \rc{t-} in \kalina, while they took on \rc{w-} in \PTir \exref{13w}.
%
The question of how this happened in detail \parencite[107--112]{meira1998proto} still has no answer, although it seems necessary to assume a scenario whereby both \rc{t-} and \rc{w-} for a time occurred on both transitive and intransitive verbs \parencite[112]{meira1998proto}.\footnote{
In fact, even the issue of \emph{when} this happened is open.
It could have happened at the \PTar stage, but the subsequent introduction of \obj{j-} in \carijo \pcref{sec:carijo} would have erased any traces of such an innovation.
Accordingly, \textcite{meira1998proto} hesitates to assign this extension to a specific proto-language.
We take a conservative stance and reconstruct it to \PTir only, but acknowledge the possibility of it happening already in \PTar.}
%
As for verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} lists the first four verbs in \cref{tab:ptirverbs} as being reconstructible to \PTar with \rc{w-}, for which we offer our reconstructed \PTir forms.
To this list, we add the other copular root \rc{eʔi} \parentext{\PTar \rc{et͡ʃi} \parencite[165]{meira1998proto}}, which has first person \obj{w-} at least in \trio.

@@

\subsection{\carijo \obj{j-}}
\label{sec:carijo}
As discussed by \textcite[105--107]{meira1998proto}, \carijo has extended the \gl{1}\gl{s_p_} marker \obj{j(i)-} to \gl{s_a_} verbs.\footnote{Since almost all \gl{s_a_} verbs are V-initial, only the \envr{}{V} allomorph \obj{j-} is attested.}
This, in combination with the extension of \gl{2}\gl{s_a_} \obj{m-} and \gl{1+2}\gl{s_a_} \obj{kɨt-}/\obj{kɨs-} to \gl{s_p_} verbs, results in a single unified \gl{s} category \exref{carpar}.

\ex<carpar> \carijo \parencites[106]{meira1998proto}[173]{robayo2000avance}\\
\begin{tabular}[t]{@{}lll@{}}
& \obj{tuda} \qu{to arrive} & \obj{eharaga} \qu{to dance}\\
\gl{1} & \obj{ji-tuda-} & \obj{j-eharaga-}\\
\gl{2} & \obj{mɨ-tuda-} & \obj{m-eharaga-}\\
\gl{1+2} & \obj{kɨsi-tuda-} & \obj{kɨs-eharaga-}\\
\gl{3} & \obj{ni-tuda-} & \obj{n-eharaga-}\\
\end{tabular}
\xe
%
Although the split-\gl{s} system has been lost entirely, former \gl{s_a_} verbs can be identified by taking a detransitivizer, like \obj{ehɨnəhɨ} \qu{to fight}, derived from \obj{hɨnəhɨ} \qu{to kill} \parencite[179]{robayo2000avance} \exref{car-30}.
%The same marker is also found in \gl{3}>\gl{1} scenarios \exref{car1p.car-1} and possessed nouns \exref{car1p.car-31}.

@@

\subsection{\yukpa \obj{j-}}
\label{sec:yukpa}
\yukpa has lost most of its \setone constructions, mostly employing innovative ones \parencite{meira2006syntactic}.
It preserves the \setone prefixes in the immediate past, where the \gl{1}\gl{s_p_} marker appears to have been extended to \gl{s_a_} verbs, as evidenced by the paradigms in \exref{yukpaintr}.
The loss of split-\gl{s} in favor of the \gl{1}\gl{s_p_} marker is accompanied by the \gl{2}\gl{s_a_} marker \obj{m(ɨ)-} displacing \gl{2}\gl{s_p_} \rc{ə(j)-}, and \gl{1+2} being lost as an inflectional value.
%It has lost the split-\gl{s} system in general

\ex<yukpaintr> \yukpa \parencites[72, 76]{largo2011yukpa}[139]{meira2006syntactic}\\
\begin{tabular}[t]{@{}llll@{}}
& \obj{otum} \qu{to wash self} & \obj{nɨ} \qu{to sleep} & \obj{ata} \qu{to fall}\\
\gl{1} & \obj{j-otum-} & \obj{jɨ-nɨ-} & \obj{j-ata-}\\
\gl{2} & \obj{m-otum-} & \obj{mɨ-nɨ-} & \obj{m-ata-}\\
\gl{3} & \obj{n-otum-} & \obj{nɨ-nɨ-} & \obj{n-ata-}\\
\end{tabular}
\xe
%
The normal intransitive first person marker is \obj{j(ɨ)-}, including those with clear reflexes of \detrz, like \obj{otum} \qu{to wash self} in \exref{yukpaintr}.
The analysis that this is a reflex of the \gl{1}\gl{s_p_} marker \rc{u(j)-} is supported by the same form occurring in \gl{3}>\gl{1} scenarios \exref{yukpatr.yuk-5}.

>>>>

\subsection{\PPek \rc{k-}}
\label{sec:pekodian}
The Pekodian branch was suggested by \textcite{meira2005southern}, as the result of fieldwork on \bakairi by Meira and the availability of more material on \ikpeng. % \parencites{pacheco1997relativizacao}{ikpengpacheco1997}{campetela1997analise}{pacheco1998nominalizacao}{ikpengpacheco2001}{campetela2002aspectos}{pacheco2002aspectos}{pacheco2003intransitivos}.
It consists of closely related \arara and \ikpeng, with \bakairi as a more distant member.
\textcite{meira2005southern} focused on phonological and lexical properties, so no reconstructive work has been done on \PPek morphosyntax.
All three Pekodian languages have a regular \gl{1}\gl{s_a_} marker \obj{k-}, as evidenced by the paradigms in \cref{tab:pekreg}.
Its presence can thus be reconstructed to \PPek.

%\ex<pekreg>
%\begin{tabular}[t]{@{}llll@{}}
%& \bakairi \qu{to go up}  & \arara \qu{to dance}  & \ikpeng \qu{to run} \\
%& \parencite[4]{meira2003bakairi} & \parencite[150]{alves2017arara} &  \parencite[52]{ikpengpacheco2001}\\
%\gl{1}\gl{s_a_} & \obj{\emp{k-}əku-} & \obj{\emp{k-}origu-} & \obj{\emp{k-}aranme-} \\
%\gl{2}\gl{s_a_} & \obj{m-əku-} & \obj{m-origu-} & \obj{m-aranme-} \\
%\gl{1+2}\gl{s_a_} & \obj{kɨd-əku-} & \obj{kud-origu-} & \obj{kw-aranme-} \\
%\gl{3}\gl{s_a_} & \obj{n-əku-} & ∅\obj{-origu-} & ∅\obj{-aranme-} \\
%\end{tabular}
%\xe

\input{floats/pekreg}

@@

The V-initial nature of \qu{to go} is evidenced in its Xinguan forms, while \bakairi \phon{\rc{wɨ}}{\obj{u}} resulted in a shift of the morpheme boundary, see also \cref{sec:go}.
The \ikpeng form \obj{aran} is made compatible with our reconstruction \rc{ɨtən} by considering that \ikpeng \obj{a} is an attested outcome of \rc{ə} (\PXin \rc{o}/\obj{e}), for example \begin{inlinelist}
 \item \obj{akari} \qu{dog} in \cref{tab:pxinw} above
 \item \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg}
 \item \obj{jaj} \qu{tree} \parencite[98]{ikpengpacheco2001} from \PC \rc{jəje}
 \end{inlinelist}.
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{\underline{ə}t\underline{ə}n}: \obj{\underline{e}r\underline{o}-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

@@

\subsection{\PWai \rc{k-}}
\label{sec:waiwaian}
This extension was briefly discussed in \cref{sec:extensions_intro}; the new \gl{1}\gl{s_p_} prefix \rc{k-}, already innovated at the \PPar stage, was later extended to \gl{1}\gl{s_a_} in \PWai.
This created a unified \gl{1}\gl{s} category, reflected in both \hixka and \waiwai \pcref{tab:pwaireg}.

\input{floats/pwaireg}

%\ex<wais> Waiwaian \rc{eɸurka} \qu{to fall} (\gl{s_a_}) and \rc{wɨnɨkɨ} \qu{to sleep} (\gl{s_p_})\\
%\begin{tabular}[t]{@{}lllllll@{}}
%& \multicolumn{2}{l}{\PWai} & \multicolumn{2}{l}{\hixka}  & \multicolumn{2}{l}{\waiwai} \\
% & \gl{s_a_} & \gl{s_p_} & \gl{s_a_} & \gl{s_p_} & \multicolumn{2}{l}{\gl{s}} \\
%\gl{1} & \rc{\emp{k-}eɸurka-} & \rc{kɨ-wɨnɨkɨ-} & \obj{\emp{k-}ehurka-} & \obj{kɨ-nɨkɨ-} & \obj{\emp{k-}eɸɨrka-} & \obj{kɨ-wɨnk-} \\
%\gl{2} & \rc{m-eɸurka-} & \rc{o-wɨnɨkɨ-} & \obj{m-ehurka-} & \obj{o-wnɨkɨ-} & \obj{m-eɸɨrka-} & \obj{mɨ-wɨnk-} \\
%\gl{1+2} & \rc{t-eɸurka-} & \rc{tɨt-wɨnɨkɨ-} & \obj{t-ehurka-} & \obj{tɨ-nɨkɨ-} & \obj{t͡ʃ-eɸɨrka-} & \obj{tɨt-wɨnk-} \\
%\gl{3} & \rc{ɲ-eɸurka-} & \rc{nɨ-wɨnɨkɨ-} & \obj{ɲ-ehurka-} & \obj{nɨ-nɨkɨ-} & \obj{ɲ-eɸɨrka-} & \obj{nɨ-wɨnk-} \\
%\end{tabular}\\
%\parencites[510]{howard2001wrought}[189--191]{hixkaryanaderby1985}[209--211]{hawkins1953waiwai}[50]{waiwaihawkins1998}
%\xe
%
\waiwai \obj{ka} \qu{to say} does not take \obj{kɨ-}, but rather conservative \obj{wɨ-} \exref{hixɨ.wai-108}.
Its \hixka counterpart also takes a reflex of \gl{1}\gl{s_a_} \rc{w-}, which has the form \obj{ɨ-} \exref{hixɨ.hix-118}.
A formally identical prefix is found on transitive verbs in \gl{1}>\gl{3} scenarios in \hixka \exref{hixɨ.hix-17}, where it also corresponds to \waiwai \obj{w(ɨ)-} \exref{hixɨ.wai-159}.

@@

\subsection{\PTir \rc{t-}}
\label{sec:taranoan}
The Taranoan languages were already grouped together by \textcite{girard1971proto}, consisting of closely related \trio and \akuriyo (subsumed here under the moniker Tiriyoan), and the more distant member \carijo.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar.
The \setone paradigms of \trio and \akuriyo contain an interesting puzzle: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.

\input{floats/ptirreg}
%This is illustrated by comparing reconstructible \PTir forms with more conservative \wayana \exref{sat} and \kalina \exref{13w}, which can be taken as representative for \PC.
%
%\ex<sat>
%\begin{tabular}[t]{@{}llllll@{}}
%\gl{1}\gl{s_a_} & \PC & \wayana & \PTir & \trio & \akuriyo\\
%\qu{yawn} & \rc{w-e-mɨta-po-} & \obj{w-emtapɨ-} & \rc{t͡ʃ-entapo-} & \obj{s-entapo-} & \obj{t͡ʃ-entapo-}\\
%\qu{bathe} & \rc{w-e-pɨ-} & \obj{w-epɨ-} & \rc{t͡ʃ-epɨ-} & \obj{s-epɨ-} & \obj{t͡ʃ-epɨ-}\\
%\qu{talk} & \rc{w-ət-uru} & \obj{w-əturu-} & \rc{t-əturu-} & \obj{t-əturu-} & \obj{k-əturu-}\footnotemark\\
%\qu{learn} & \rc{w-ət-eme-pa} & \obj{w-əhepa-} & \rc{t-əempa-}\footnotemark & \obj{t-əenpa-} & \obj{k-əempa-}\\
%\end{tabular}\\
%\parencites[28]{camargo2010wayana}[292, 664]{triomeira1999}[77, 86, 82]{gildea1994akuriyo}[50, 208, 213]{wayanatavares2005}
%\xe
%\addtocounter{footnote}{-1}
%\footnotetext{\PTir \rc{t-} was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.}
%\addtocounter{footnote}{1}
%\footnotetext{
%The \trio form is not due to dissimilation from \rc{mp} to \obj{np}, but rather a phonemic transcription rendering the syllable-final archiphoneme \pnm{N} as \obj{n} \parencite[31]{triomeira1999}.
%Further, \PTar intervocalic \rc{t͡ʃ} was lost entirely in \trio (except after \rc{i}) and is reflected as \obj{ʔ} in \akuriyo \parencite[31--32]{meira1998proto}.
%Given the original shape of the detransitivizers \rc{ət(e)-} and \rc{e-}, \trio \obj{ə-enpa} \qu{to learn} contains a reflex of the former with loss of \rc{t}, so \akuriyo \obj{əempa} would be expected to have a glottal stop: ?\obj{əʔempa}.
%Its absence could either be due to \trio influence or not being transcribed by \textcite{gildea1994akuriyo}.
%Since there is no evidence for a glottal stop in \PTir (\rc{əʔempa}), I do not reconstruct it.}
%
%\ex<13w> 
%\begin{tabular}[t]{@{}llllll@{}}
%\gl{1}>\gl{3}& \PC & \kalina & \PTir & \trio & \akuriyo \\
%\qu{see} & \rc{t-ene-} & \obj{s-ene-} & \rc{w-ene-} & \obj{w-ene-} & \obj{(w-)ene-} \\
%\qu{bring} & \rc{t-enepɨ-} & \obj{s-enepɨ-} & \rc{w-enepɨ-} & \obj{w-enepɨ-} & Ø\obj{-enepɨ-} \\
%\end{tabular}\\
%\parencites[425--426]{hoff1968carib}[28, 304]{triomeira1999}[86--87]{gildea1994akuriyo}
%\xe
%%
%Thus, \wayana preserves the \PC \gl{1}\gl{s_a_} marker \rc{w-}, while it was replaced by \rc{t-} (\phonr{}{\rc{t͡ʃ-}}{\rc{e}}) in \PTir \exref{sat}.
%%Both \qu{to yawn} and \qu{to bathe} are transparently derived \gl{s_a_} verbs, from \rc{mɨta-po} \qu{to open the mouth of} and \rc{pɨ} \qu{to wash}, respectively.
%On the other hand, \gl{1}>\gl{3} scenarios show a reflex of \rc{t-} in \kalina, while they took on \rc{w-} in \PTir \exref{13w}.
%
The question of how this happened in detail \parencite[107--112]{meira1998proto} still has no answer, although it seems necessary to assume a scenario whereby both \rc{t-} and \rc{w-} for a time occurred on both transitive and intransitive verbs \parencite[112]{meira1998proto}.\footnote{
In fact, even the issue of \emph{when} this happened is open.
It could have happened at the \PTar stage, but the subsequent introduction of \obj{j-} in \carijo \pcref{sec:carijo} would have erased any traces of such an innovation.
Accordingly, \textcite{meira1998proto} hesitates to assign this extension to a specific proto-language.
We take a conservative stance and reconstruct it to \PTir only, but acknowledge the possibility of it happening already in \PTar.}
%
As for verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} lists the first four verbs in \cref{tab:ptirverbs} as being reconstructible to \PTar with \rc{w-}, for which we offer our reconstructed \PTir forms.
To this list, we add the other copular root \rc{eʔi} \parentext{\PTar \rc{et͡ʃi} \parencite[165]{meira1998proto}}, which has first person \obj{w-} at least in \trio.

@@

\subsection{\carijo \obj{j-}}
\label{sec:carijo}
As discussed by \textcite[105--107]{meira1998proto}, \carijo has extended the \gl{1}\gl{s_p_} marker \obj{j(i)-} to \gl{s_a_} verbs.\footnote{Since almost all \gl{s_a_} verbs are V-initial, only the \envr{}{V} allomorph \obj{j-} is attested.}
This, in combination with the extension of \gl{2}\gl{s_a_} \obj{m-} and \gl{1+2}\gl{s_a_} \obj{kɨt-}/\obj{kɨs-} to \gl{s_p_} verbs, results in a single unified \gl{s} category \pcref{tab:carreg}.
%
\input{floats/carreg}
%
Although the split-\gl{s} system has been lost entirely, former \gl{s_a_} verbs can be identified by taking a detransitivizer, like \obj{ehɨnəhɨ} \qu{to fight}, derived from \obj{hɨnəhɨ} \qu{to kill} \parencite[179]{robayo2000avance} \exref{car-30}.
%The same marker is also found in \gl{3}>\gl{1} scenarios \exref{car1p.car-1} and possessed nouns \exref{car1p.car-31}.

@@

\subsection{\yukpa \obj{j-}}
\label{sec:yukpa}
\yukpa has lost most of its \setone constructions, mostly employing innovative ones \parencite{meira2006syntactic}.
It preserves the \setone prefixes in the immediate past, where the \gl{1}\gl{s_p_} marker appears to have been extended to \gl{s_a_} verbs, as evidenced by the paradigms in \cref{tab:yukreg}.
The loss of split-\gl{s} in favor of the \gl{1}\gl{s_p_} marker is accompanied by the \gl{2}\gl{s_a_} marker \obj{m(ɨ)-} displacing \gl{2}\gl{s_p_} \rc{ə(j)-}, and \gl{1+2} being lost as an inflectional value.
%It has lost the split-\gl{s} system in general

%\ex<yukpaintr> \yukpa \parencites[72, 76]{largo2011yukpa}[139]{meira2006syntactic}\\
%\begin{tabular}[t]{@{}llll@{}}
%& \obj{otum} \qu{to wash self} & \obj{nɨ} \qu{to sleep} & \obj{ata} \qu{to fall}\\
%\gl{1} & \obj{j-otum-} & \obj{jɨ-nɨ-} & \obj{j-ata-}\\
%\gl{2} & \obj{m-otum-} & \obj{mɨ-nɨ-} & \obj{m-ata-}\\
%\gl{3} & \obj{n-otum-} & \obj{nɨ-nɨ-} & \obj{n-ata-}\\
%\end{tabular}
%\xe

\input{floats/yukreg}

The normal intransitive first person marker is \obj{j(ɨ)-}, including those with clear reflexes of \detrz, like \obj{otum} \qu{to wash self} in \cref{tab:yukreg}.
The analysis that this is a reflex of the \gl{1}\gl{s_p_} marker \rc{u(j)-} is supported by the same form occurring in \gl{3}>\gl{1} scenarios \exref{yukpatr.yuk-5}.
